---
Author: Patrick Ho
Category: Programming
Output: md_document
Slug: r-data-types
Tags: R
Title: Data Types
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

## Jump to:
- [Objects](#objects)
- [Vectors](#vectors)
- [Matrices and Arrays](#matrices_arrays)
- [Factors](#factors)
- [Lists](#lists)
- [Data Frames](#data_frames)
- [str](#str)

## Objects {#objects}
- Five atomic classes of objects: **Numeric, Integer, Character, Logical, Complex**
- Numbers are treated as double precision by default
    - **Inf** is infinity, e.g. 1/0
    - **NaN** is not a number (undefined), e.g. 0/0
- Objects can have **attributes**
    - names, dimensions, class, length
    - user-defined metadata

## Vectors {#vectors}
- All elements must be same type

```{r vectors, eval=FALSE}
# Create
x <- c(1, 5.5, 2.7, 3)     # Numeric - Defaults to double
x <- c(3L, 8L, 22L, 2L)    # Integer - Use L suffix
x <- 1:4                   # Integer - Range defaults to integer
x <- c(TRUE, FALSE, T, F)  # Logical
x <- c('a', 'b', 'c')      # Character
x <- c(1+0i, 2+4i)         # Complex
x <- vector('numeric', length = 5)  # Numeric - Default values of 0

# Inspect
typeof(x)  # E.g. "character"
is.integer(x)
is.double(x)
is.numeric(x)  # TRUE if either integer or numeric
is.logical(x)
is.character(x)

# Check missing
x <- c(1, 2, NaN, NA, 4)
is.na(x)   # FALSE FALSE  TRUE  TRUE FALSE - NaN is also NA
is.nan(x)  # FALSE FALSE  TRUE FALSE FALSE

# Coercion
x <- 0:3
as.numeric(x)    # 0 1 2 3
as.character(x)  # "0" "1" "2" "3"
as.logical(x)    # FALSE TRUE TRUE TRUE - FALSE if value is 0, TRUE otherwise

x <- c('abc', 123)  # c('abc', '123') - coerced so all elements are same type

x <- c('a','b')
as.numeric(x)    # NA NA - NAs introduced by coercion due to invalid conversion

# Vectorized operations
x <- 1:3; y <- 4:6
x + y   # 5 7 9
x >= 2  # F T T
x == 3  # F F T
x * y   # 4 10 18
x / y   # 0.25 0.40 0.50

# Summarise
x <- c(1, 2, 3)
length(x)  # 3
sum(x)     # 6
mean(x)    # 2
```

## Matrices and Arrays {#matrices_arrays}
- Vectors with a dimension attribute will become multi-dimensional *array*
- An array with 2 dimensions is a *matrix*
- Can also apply dimensions to a list

```{r matrices_arrays, eval=FALSE}
# Create from vector
x <- 1:6
dim(x) <- c(2,3)  # Turn vector into 2x3 matrix - Column-wise by default "top-left then down"
matrix(1:6, nrow = 2, ncol = 3)  # Same result
matrix(1:6, c(2,3))  # Same result

# Create using cbind/rbind
x <- 1:3
y <- 4:6
cbind(x, y)  # 3x2 matrix - bind vectors as columns
rbind(x, y)  # 2x3 matrix - bind vectors as rows

# Inspect
ncol(x)  # Number of columns
nrow(x)  # Number of rows

# Manipulate
rownames(x) <- c('r1', 'r2')  # Set matrix row names
colnames(x) <- c('c1', 'c2', 'c3')  # Set matrix col names
t(x)  # Transpose matrix

# Vectorized operations
x <- matrix(1:4, 2, 2); y <- matrix(5:8, 2, 2)
x * y    # element-wise multiplication
x / y    # element-wise division
x %*% y  # matrix multiplication

# Array
y <- array(1:12, c(2,2,3))  # 3-dimensional array
dimnames(y) <- list(c('a1', 'a2'), c('b1', 'b2'), c('c1', 'c2', 'c3'))  # Set array dimension names
```

## Factors {#factors}
- Vectors that can only contain predefined values (levels)
- Use for categorical data

```{r factors, eval=FALSE}
# Create
x <- factor(c('m', 'f', 'f', 'm', 'm'))

# Inspect
is.factor(x)
levels(x)   # "f" "m" are the only valid values
table(x)    # 2 f, 3 m

# Define levels
x[1] <- 'a'  # Invalid factor level, NA generated
x <- factor(x, levels = c('m', 'f', 'a'))  # Define levels, m=1 f=2 a=3
x[1] <- 'a'  # Will now work as 'a' is a valid factor level
```

## Lists {#lists}
- Elements can be any type, even nested lists

```{r lists, eval=FALSE}
# Create
x <- list(c(5.5, 2.7), 3L:8L, c(T, F, T), c('abc', 'def'), list(1, 'a', F))

# Inspect
is.list(x)

# Coercion
as.list(c(1, 2, 3))    # list(1, 2, 3)
unlist(list(1, 2, 3))  # c(1, 2, 3), coerce to atomic vector

# Manipulate
names(x) <- c('num', 'int', 'logi', 'char', 'list')
```

## Data Frames {#data_frames}
- List of equal length vectors
- Shares same properties as matrix and list

```{r data_frames, eval=FALSE}
# Create
df <- data.frame(col1 = 1:3, col2 = c('a', 'b', 'c'))
df <- data.frame(col1 = 1:3,col2 = c('a', 'b', 'c'),
                 stringsAsFactors = FALSE)  # Leave col2 as chr

# Inspect
is.data.frame(df)
ncol(x)  # Number of columns
nrow(x)  # Number of rows

# Manipulate
cbind(df, data.frame(col3 = c(T, F, T)))  # Add column
rbind(df, data.frame(col1 = 4, col2 = 'd'))  # Add row
names(df) <- c('col_a', 'col_b')   # Set names of columns
```

## str {#str}
- Display internal structure of an R object in a compact way

```{r str, eval=FALSE}
# Vectors
x <- runif(100)
summary(x)  # Quartiles
str(x)  # num [1:100] 0.3257 0.3514 0.0628 0.2261 0.0838 ...

# Matrix
x <- matrix(1:100, 10, 10)
x[, 1]  # [1]  1  2  3  4  5  6  7  8  9 10
str(x)  # int [1:10, 1:10] 1 2 3 4 5 6 7 8 9 10 ...

# Factors
x <- gl(3, 5)
summary(x)  # Frequency counts
str(x)  # Factor w/ 3 levels "1","2","3": 1 1 1 1 1 2 2 2 2 2 ...

# Lists
x <- list(c(5.5, 2.7), list(1, 'a', F))
str(x)
# List of 2
#  $ : num [1:2] 5.5 2.7
#  $ :List of 3
#   ..$ : num 1
#   ..$ : chr "a"
#   ..$ : logi FALSE

# Data Frames
str(airquality)
# 'data.frame':	153 obs. of  6 variables:
#  $ Ozone  : int  41 36 12 18 NA 28 23 19 8 NA ...
#  $ Solar.R: int  190 118 149 313 NA NA 299 99 19 194 ...
```
