---
Author: Patrick Ho
Category: Programming
Date: 2017-02-12
Slug: r-files
Summary:
Tags: R
Title: Files
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

## Tabular Data {#tabular}

```{r tabular, eval=FALSE}
# Read
infile <- 'input.csv'
df <- read.table(file = infile,
                 header = TRUE,  # Has header line?
                 sep = ',',  # Separator character
                 nrows = 100,  # Number of rows to read
                 comment.char = '#',  # Skip rows starting with this character
                 skip = 2, # Skip this many rows from start
                 stringsAsFactors = TRUE)  # Encode chars as factors
df <- read.csv(infile)  # Same as read.table() with comma as default separator


# Read large files
sample <- read.table('large_file.csv', nrows = 100)
classes <- sapply(sample, class)
df <- read.table('large_file.csv',
                 colClasses = classes,  # Pre-defining column classes is much faster
                 nrows = 50000)  # Pre-defining rows (Unix wc) can help with memory usage


# Write
outfile <- 'output.csv'
write.table(df,
            file = outfile,
            append = FALSE)  # FALSE will overwrite
write.csv(df, outfile)
```

## R Objects {#objects}
- Text format and includes meta-data
- Easy to edit but less space efficient
- Works nicely with version control programs like git
- **dget** pairs with **dput**
- **source** pairs with **dump**

```{r code, eval=FALSE}
# Read
dget('data.R')

source('data.R')

# Write
df <- data.frame(a = 1, b = 'b')
dput(df)  # Prints out structure
dput(df, 'data.R')

x <- 5
y <- 'abc'
dump(c('x','y','df'), file = 'data.R')
```

## Connections {#connections}
#### Interfaces
- file()
- gzfile() = connection to gzip compressed file
- url() = connection to webpage

#### Modes
- 'r' = read only
- 'w' = writing/creating
- 'a' = appending
- 'rb', 'wb', 'ab' = read/write/append in binary mode (Windows)

```{r connections, eval=FALSE}
# Read
con <- file('input.csv', 'r')
con <- gzfile('input.gz', 'r')
con <- url('http://cran.r-project.org/', 'r')
x <- readLines(con, 10)  # Read in first 10 lines
close(con)
```

## Workspace {#workspace}

```{r workspace, eval=FALSE}
# Read
load('mydata.RData')

# Write
save.image(file = 'mydata.Rdata')  # Defaults to '.Rdata' if not specified
```
