---
Author: Patrick Ho
Category: Programming
Date: 2017-02-12
Slug: r-scoping-rules
Summary:
Tags: R
Title: Scoping Rules
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```
## Namespaces
R searches for a symbol in the following order:

1. Search the global environment
    - User's workspace, always searched first
2. Search the namespaces of each package on the search list
    - Ordering matters
    - Most recently loaded package goes in position 2 and everything else shifts down
    - Base package is always last

Separate namespaces for functions and non-functions, e.g. can have both 'x' function and 'x' object.

```{r namespaces}
# Search list
search()
```

## Lexical Scoping
- R uses *Lexical Scoping* or *Static Scoping*
    - As opposed to *Dynamic Scoping*
- Determines how values are assigned to *free variables*
- Values of free variables are searched for in the environment in which the function as defined
    - If value not found, search is continued in the *parent environment*
    - Continues upwards until *top-level environment*, usually global environment or package namespace
    - Then continues downwards until *empty environment*
    - If still not found, throw error

```{r lexical_scoping, eval=FALSE}
# Function below has formal arguments x/y, and free variable z
f <- function(x, y) {
  x + y - z
}


# Example: Function generator - Value of free variable n taken from function definition environment
make_power_fn <- function(n) {
  pow <- function(x) {
    x^n
  }
  pow
}

cube_fn <- make_power_fn(3)
cube_fn(3)                      # 27
square_fn <- make_power_fn(2)
square_fn(3)                    # 9

# Explore function environment
ls(environment(cube_fn))        # "n"   "pow"
get('n', environment(cube_fn))  # 3
ls(environment(square_fn))      # "n"   "pow"
get('n', environment(square_fn))  # 2


# Example: Lexical (value from definition env) vs Dynamic Scoping (value from calling env)
y <- 10

f <- function(x) {
  y <- 2
  y^2 + g(x)
}

g <- function(x) {
  x*y
}

f(3)  # 34 because y=10 in environment where g() is defined, so (2^2 + 3*10)
```
