---
Author: Patrick Ho
Category: Programming
Slug: r-simulation
Tags: R
Title: Simulation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

## Distributions {#distributions}
Probability distributions are already built in:
- **norm** for Normal
- **binom** for Binomial
- **pois** for Poisson
- etc.

Each distribution has four functions with the following prefix:
- **r** for random, random numbers from distribution
- **d** for density, density function (PDF)
- **p** for probability, cumulative distribution function (CDF)
- **q** for quantile, inverse CDF

```{r distributions, eval=FALSE}
# Normal
rnorm(10)   # 1.51178117  0.38984324 -0.62124058 -2.21469989 ...
pnorm(0.5)  # 0.6914625  <- 69% prob that value is <=0.5
qnorm(0.6914625)  # 0.5  <- Inverse of above

# Binomial
rbinom(10, 5, 0.5)  # 5 3 2 2 1 0 3 1 2 3  <- 10 obs, 5 trials each, 50% prob per trial
dbinom(3, 5, 0.5)  # 0.3125  <- 31.25% prob of getting exactly 3 success in 5 trials at 50% prob
```

# Seed {#seed}

```{r seed, eval=FALSE}
set.seed(1)
rnorm(5)  # [1] -0.6264538  0.1836433 -0.8356286  1.5952808  0.3295078
rnorm(5)  # [1] -0.8204684  0.4874291  0.7383247  0.5757814 -0.3053884

set.seed(1)
rnorm(5)  # [1] -0.6264538  0.1836433 -0.8356286  1.5952808  0.3295078  <- Same as first
```

# Sampling {#sampling}

```{r sampling, eval=FALSE}
set.seed(1)

# Random sample
x <- 1:9
sample(x, 3)  # 3 9 5
sample(x, 3)  # 9 2 7  <- New sample with each call
sample(letters, 5)  # y" "q" "p" "b" "e"

# Random sample with replacement
sample(x, 5, replace = TRUE)  # 2 7 4 7 5

# Permutations
sample(x)  # 7 8 3 5 6 1 2 4 9
sample(x)  # 4 1 3 6 2 5 9 8 7  <- New permutation with each call
```

# Linear Model {#linear_model}

```{r linear_model, eval=TRUE}
set.seed(1)

x <- rnorm(100)
e <- rnorm(100, 0, 0.5)  # Add noise with mean 0, sd 0.5
y <- 3 + 2 * x + e

plot(x, y)
```
